---
title: "SICK_FigS2: Unenriched_regression"
author: "M. Foster"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries
```{r}
library(tidyverse)
library(broom)
library(multcomp)
library(dplyr)
library(RColorBrewer)

# Set the theme for ggplot2
theme_set(theme_minimal())
```

## Load Data
```{r}
# Load datasets
prospective_data <- read.csv("Day0_clinical_data.csv") %>% as_tibble()
unenriched_data <- read.csv("unenriched_MOFA.csv") %>% as_tibble()

```
#clean up data
```{r}

# Filter data
filtered_data <- prospective_data %>%
  dplyr::select(Subject.ID, Sample.ID) # Select PID and the first DukeID column

# Clean the data: Keep Accession column and fix column names
cleaned_unenriched_data <- unenriched_data %>%
  dplyr::rename_with(~ str_remove(., "^X"), -Accession)  # Remove leading X from column names except Accession

# Create a logical vector indicating which columns in `cleaned_unenriched_data` match `DukeID` in `data`
matched_columns <- colnames(cleaned_unenriched_data) %in% filtered_data$Sample.ID

# Keep only the `Accession` column and matched columns, then rename them with `PID`
updated_unenriched_data <- cleaned_unenriched_data %>%
  dplyr::select(Accession, all_of(colnames(cleaned_unenriched_data)[matched_columns])) %>%  # Filter for matched columns
  rename_with(
    ~ ifelse(. %in% filtered_data$Sample.ID, filtered_data$`Subject.ID`[match(., filtered_data$Sample.ID)], .), 
    -Accession # Exclude `Accession` from renaming
  )


# Display the filtered data
head(updated_unenriched_data)

```
##knn imputation
```{r}

library(VIM)


# Perform KNN imputation
imputed_unenriched_data <- kNN(updated_unenriched_data, k = 5, variable = names(updated_unenriched_data)[-1])  # Replace 'k' with the desired number of neighbors

# Remove the `.imp` columns added by kNN
imputed_unenriched_data <- imputed_unenriched_data[, !grepl("\\imp$", names(imputed_unenriched_data))]

# Optional: Display summary of imputed data
summary(imputed_unenriched_data)

```

## Log2 Transformation of unenriched Data
```{r}
imputedlog2unenriched_data <- imputed_unenriched_data %>%
  mutate(across(-Accession, ~log2(. + 1)))  # Apply log2 transformation to all columns except 'Accession'

# Display a preview of the transformed data
head(imputedlog2unenriched_data)

# Save the transformed MOFA data
write.csv(imputedlog2unenriched_data, "Log2_Transformed_unenriched_Data.csv", row.names = FALSE)
```

## Align Data and Confirm Matching IDs
```{r}

# Standardize column names in data to match `Subject.ID` in prospective data
colnames(imputedlog2unenriched_data) <- colnames(imputedlog2unenriched_data) %>%
  str_replace_all("\\.", "-")  # Replace `.` with `-`

# Check if all `` values in prospective_data are in column names
missing_Subject.ID <- setdiff(prospective_data$Subject.ID, colnames(imputedlog2unenriched_data))
if (length(missing_Subject.ID) > 0) {
  print("The following Subject.IDs are missing in unenriched data:")
  print(missing_Subject.ID)
} else {
  print("All Subject.IDs in prospective data are matched in unenriched data.")
}

# Align unenriched data with filtered prospective data
aligned_unenriched <- imputedlog2unenriched_data %>%
  dplyr::select(any_of(c("Accession", prospective_data$Subject.ID)))  # Ensure Accession is always included

# Save the aligned MOFA data
write.csv(aligned_unenriched, "Aligned_unenriched_Data.csv", row.names = FALSE)
write.csv(prospective_data, "Prospective_data.csv", row.names = FALSE)
```
#perform regression on all dependent varoiables
```{r}

# List all columns except the first two
predictor_columns <- colnames(prospective_data)[-c(1, 2)]

# Loop through each dependent variable in predictor_columns
for (dependent_variable in predictor_columns) {
  
  # Step 1: Filter the data for the current dependent variable (remove rows with missing values)
  data_filtered <- prospective_data %>%
    dplyr::select(Subject.ID, all_of(dependent_variable)) %>%
    drop_na()
  
  # Step 2: Reshape MOFA data and assign the dependent variable values to the 4th column
  aligned_unenriched_long <- aligned_unenriched %>%
    pivot_longer(cols = -Accession, names_to = "Subject.ID", values_to = "independent_value") %>%
    inner_join(data_filtered, by = "Subject.ID") %>%
    # Assign the dependent variable values to the 4th column ('dependent_value')
    mutate(dependent_value = .[[dependent_variable]]) %>%
    # Select only the necessary columns: Accession, Subject.ID, independent_value, dependent_value
    dplyr::select(Accession, Subject.ID, independent_value, dependent_value)

  # Step 3: Perform regression using the dependent and independent variable for each Accession
  regression_results <- aligned_unenriched_long %>%
    group_by(Accession) %>%
    do({
      # Apply regression model for each Accession group
      model <- lm(dependent_value ~ independent_value, data = .)
      
      # Get the model summary and extract the relevant coefficients, p-values, and R-squared
      data.frame(
        Coefficient = coef(model)[2],  # Get the coefficient of the independent variable
        P_Value = summary(model)$coefficients[2, 4],  # Get the p-value of the independent variable
        R_Squared = summary(model)$r.squared  # Get R-squared value
      )
    }) %>%
    ungroup() %>%
    mutate(Q_Value = p.adjust(P_Value, method = "fdr"))  # Apply FDR correction for p-values

  # Step 4: Print regression results for the current dependent variable
  print(regression_results)

  # Step 5: Save the results to a CSV file with a dynamic filename based on the dependent variable
  filename <- paste0("regression_results_", dependent_variable, ".csv")
  write.csv(regression_results, filename, row.names = FALSE)
  cat("Results saved to", filename, "\n")
  
  # Step 6: Pause after processing each dependent variable
  #readline(prompt = paste("Press Enter to continue to the next dependent variable (", dependent_variable, ")..."))
}

# Now, the regression results for each dependent variable are saved with appropriate filenames.
```
#combine regression
```{r}

#collect regression data
csv_files <- list.files(pattern = "^regression_results_.*\\.csv$", full.names = TRUE)

# Initialize an empty list to store each data frame
all_data_frames <- list()

# Loop over the CSV files, load them, and add to the list
for (file in csv_files) {
  
  # Load the CSV file
  df <- read.csv(file)
  
  # Check if 'Accession' column exists
  if (!"Accession" %in% colnames(df)) {
    stop(paste("Error: 'Accession' column not found in file:", file))
  }
  
  # Standardize the 'Accession' column header by removing everything after the first underscore
  colnames(df)[colnames(df) == "Accession"] <- sub("_.*", "", "Accession")
  
  # Extract the dependent variable from the filename (e.g., 'age' from 'regression_results_age.csv')
  dependent_variable <- gsub("regression_results_(.*).csv", "\\1", basename(file))
  
  # Rename the regression result columns (e.g., Coefficient, P_Value) to reflect the dependent variable (e.g., Coefficient_age, P_Value_age)
  colnames(df)[!colnames(df) %in% "Accession"] <- paste(colnames(df)[!colnames(df) %in% "Accession"], dependent_variable, sep = "_")
  
  # Add the data frame to the list
  all_data_frames[[dependent_variable]] <- df
}

# Combine all data frames into one by aligning them based on the 'Accession' column
combined_data <- Reduce(function(x, y) merge(x, y, by = "Accession", all = TRUE), all_data_frames)

# View the combined data
head(combined_data)

# Save the combined results to a new CSV file
write.csv(combined_data, "combined_regression_results.csv", row.names = FALSE)
cat("Combined results saved to combined_regression_results.csv\n")

```
#import crp regression results

```{r}
# Import the CSV file
crp_data <- read_csv("regression_results_crp.csv")

# Display the first few rows of the dataset
head(crp_data)

```
#filter crp data for lollipop plot

```{r}

# Filter the data for Q-value < 0.01 and sort by Q-value (low to high)
filtered_crp_data <- crp_data %>%
  filter(Q_Value < 0.01) %>%
  #filter(R_Squared > 0.47) %>%
  arrange(Q_Value)

# Remove the "_HUMAN" suffix for Accession
filtered_crp_data$Accession <- gsub("_HUMAN", "", filtered_crp_data$Accession)

# Convert Accession to a factor with levels ordered by Q-value (lowest Q-value at the top)
filtered_crp_data$Accession <- factor(filtered_crp_data$Accession, 
                                  levels = filtered_crp_data$Accession[order(filtered_crp_data$Q_Value)])

# Reverse the factor levels so that the smallest Q-value is at the top
filtered_crp_data$Accession <- factor(filtered_crp_data$Accession, levels = rev(levels(filtered_crp_data$Accession)))


```

#CRP lollipop plot

```{r}
# Create the lollipop plot using ggplot2 with reversed YlGnBu palette
lollipop_plot <- ggplot(filtered_crp_data, aes(x = Coefficient, y = Accession)) +
  # Add vertical lines representing the coefficient values (black lines)
  geom_segment(aes(x = 0, xend = Coefficient, y = Accession, yend = Accession), color = "black", size = 1) +
  # Add circles at the end of each line, with size proportional to R-squared
  geom_point(aes(color = Q_Value, size = R_Squared), shape = 16) +
  scale_color_gradientn(colors = rev(brewer.pal(9, "YlGnBu")[3:9]), name = "Q-value") +  # Reverse color scale (dark to light)
  
  # Manually define the breaks for the size legend based on R-squared range
  scale_size_continuous(
    range = c(3, 10),  # Control size range of the points
    name = "R-squared",
    breaks = seq(from = min(filtered_crp_data$R_Squared, na.rm = TRUE), 
                 to = max(filtered_crp_data$R_Squared, na.rm = TRUE), 
                 length.out = 7),  # More breaks (7 here) to create more size guides
    labels = scales::number_format()(seq(from = min(filtered_crp_data$R_Squared, na.rm = TRUE), 
                                         to = max(filtered_crp_data$R_Squared, na.rm = TRUE), 
                                         length.out = 7))  # Format the labels based on breaks
  ) +
  
  labs(x = "Coefficient", y = "Protein Group") +  # Removed title
  theme(axis.text.x = element_text(face = "bold", angle = 45, hjust = 1, size = 16),  # Increase x-axis font size
        axis.text.y = element_text(face = "bold", size = 16),  # Increase y-axis font size and make bold
        axis.title.x = element_text(face = "bold", size = 20),  # Increase x-axis title font size
        axis.title.y = element_text(face = "bold", size = 20),  # Increase y-axis title font size
        plot.title = element_text(size = 20, face = "bold")) +  # Increase plot title font size
  
  coord_cartesian(clip = "off") +  # Allow top circle to extend beyond the plot area
  guides(
    size = guide_legend(reverse = TRUE, 
                        title = "R-squared",
                        labels = scales::number_format()(seq(from = min(filtered_crp_data$R_Squared, na.rm = TRUE), 
                                                              to = max(filtered_crp_data$R_Squared, na.rm = TRUE), 
                                                              length.out = 7))),  # Format labels in the legend
    color = guide_colorbar(reverse = TRUE)  # Reverse the Q-value legend (lowest Q-value at the top)
  )

# Save the plot as a PNG file
ggsave("crp_lollipop_plot.png", plot = lollipop_plot, device = "png", dpi = 600)

```

#import wbc regression results

```{r}
# Import the CSV file
wbc_data <- read_csv("regression_results_wbc.csv")

# Display the first few rows of the dataset
head(wbc_data)

```
#filter wbc data for lollipop plot

```{r}

# Filter the data for Q-value < 0.01 and sort by Q-value (low to high)
filtered_wbc_data <- wbc_data %>%
  filter(Q_Value < 0.01) %>%
  filter(R_Squared > 0.55) %>%
  arrange(Q_Value)

# Remove the "_HUMAN" suffix for Accession
filtered_wbc_data$Accession <- gsub("_HUMAN", "", filtered_wbc_data$Accession)

# Convert Accession to a factor with levels ordered by Q-value (lowest Q-value at the top)
filtered_wbc_data$Accession <- factor(filtered_wbc_data$Accession, 
                                  levels = filtered_wbc_data$Accession[order(filtered_wbc_data$Q_Value)])

# Reverse the factor levels so that the smallest Q-value is at the top
filtered_wbc_data$Accession <- factor(filtered_wbc_data$Accession, levels = rev(levels(filtered_wbc_data$Accession)))


```

#wbc lollipop plot

```{r}
# Create the lollipop plot using ggplot2 with reversed YlGnBu palette
lollipop_plot <- ggplot(filtered_wbc_data, aes(x = Coefficient, y = Accession)) +
  # Add vertical lines representing the coefficient values (black lines)
  geom_segment(aes(x = 0, xend = Coefficient, y = Accession, yend = Accession), color = "black", size = 1) +
  # Add circles at the end of each line, with size proportional to R-squared
  geom_point(aes(color = Q_Value, size = R_Squared), shape = 16) +
  scale_color_gradientn(colors = rev(brewer.pal(9, "YlGnBu")[3:9]), name = "Q-value") +  # Reverse color scale (dark to light)
  
  # Manually define the breaks for the size legend based on R-squared range
  scale_size_continuous(
    range = c(3, 10),  # Control size range of the points
    name = "R-squared",
    breaks = seq(from = min(filtered_wbc_data$R_Squared, na.rm = TRUE), 
                 to = max(filtered_wbc_data$R_Squared, na.rm = TRUE), 
                 length.out = 7),  # More breaks (7 here) to create more size guides
    labels = scales::number_format()(seq(from = min(filtered_wbc_data$R_Squared, na.rm = TRUE), 
                                         to = max(filtered_wbc_data$R_Squared, na.rm = TRUE), 
                                         length.out = 7))  # Format the labels based on breaks
  ) +
  
  labs(x = "Coefficient", y = "Protein Group") +  # Removed title
  theme(axis.text.x = element_text(face = "bold", angle = 45, hjust = 1, size = 16),  # Increase x-axis font size
        axis.text.y = element_text(face = "bold", size = 16),  # Increase y-axis font size and make bold
        axis.title.x = element_text(face = "bold", size = 20),  # Increase x-axis title font size
        axis.title.y = element_text(face = "bold", size = 20),  # Increase y-axis title font size
        plot.title = element_text(size = 20, face = "bold")) +  # Increase plot title font size
  
  coord_cartesian(clip = "off") +  # Allow top circle to extend beyond the plot area
  guides(
    size = guide_legend(reverse = TRUE, 
                        title = "R-squared",
                        labels = scales::number_format()(seq(from = min(filtered_wbc_data$R_Squared, na.rm = TRUE), 
                                                              to = max(filtered_wbc_data$R_Squared, na.rm = TRUE), 
                                                              length.out = 7))),  # Format labels in the legend
    color = guide_colorbar(reverse = TRUE)  # Reverse the Q-value legend (lowest Q-value at the top)
  )

# Save the plot as a PNG file
ggsave("wbc_lollipop_plot.png", plot = lollipop_plot, device = "png", dpi = 600)

```


